---
title: "Module 4 - Tidying Data with Tidyverse"
author: "Lansing Perng"
date: "Fall 2023"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

Tidyverse is a collection of R packages that are designed to work together seamlessly to make data manipulation, visualization, and analysis easier and more intuitive. In tidy data, each variable is a column and each observation is a row. This makes it easy to use functions like dplyr and ggplot2 to manipulate and visualize data.

Some of the most popular packages in Tidyverse include:

- `readr`: for reading in and parsing various file types.
- `dplyr`: for data manipulation tasks such as filtering, selecting, and summarizing data.
- `tidyr`: for reshaping and tidying data.
- `stringr`: for string manipulation tasks.
- `lubridate`: for working with dates and times.
- `ggplot2`: for creating high-quality, customizable visualizations.
- `purrr`: for functional programming tasks such as mapping and reducing.

One of the main advantages of Tidyverse is that it provides a consistent and intuitive syntax across packages, making it easier to learn and use. It also emphasizes the use of piping (%>%) to chain functions together, which can make code more readable and efficient.

Overall, Tidyverse has become a popular choice for data scientists and analysts due to its powerful and user-friendly tools for data manipulation and visualization. You can use tidyverse outputs with packages outside of the tidyverse, but sometimes this may require minor manipulation or conversion. Basically, tidyverse is the Apple of the R world.

```{r}
######## tidyr & stringr

## recoding data (replacing character names in one column with a code)
# Example data
data <- data.frame(sex = c("M", "F", "M", "F", "M"))

# Recode the sex column using tidyverse
data$scode <- recode(data$sex, "M" = "1", "F" = "2")

# View the updated data
data

## some less common but useful functions
library(tidyr)

# create sample data frame
df <- data.frame(
  name = c("Alice", "Bob", "Charlie"),
  age = c(30, NA, 25),
  city = c("New York, Los Angeles", "Chicago", "Seattle")
)
print(df)

# drop rows with missing values
df_drop <- drop_na(df)
print(df_drop)

# replace missing values with another specified value
df_replace <- replace_na(df, list(age = 0))
print(df_replace)


##### stringr

library(stringr)

# Create a string
my_string <- "The quick brown fox jumps over the lazy dog"

# Get the length of the string
str_length(my_string)

# Extract a substring
str_sub(my_string, start = 5, end = 9)

# Replace a substring
str_replace(my_string, "brown", "red")

# Check if a pattern is present
str_detect(my_string, "fox")

# Extract a pattern
str_extract(my_string, "brown|red") #brown OR red

# Split a string
vec<- unlist(str_split(my_string, " ")) #function creates list, which can be incompatible with certain functions
#unlisting turns it into a workable vector
df1<- data.frame(vec)

#demo separate()
# create example data frame
df <- data.frame(
  cities1 = c("New York, NY", "Los Angeles, CA", "Chicago, IL", "Miami, FL")
)
# separate columns at comma
df_sep <- separate(df, cities1, into = c("city1", "state1"), sep = ",\\s*")


```



```{r}
# Load lubridate library
library(lubridate)

# Create a date object
date1 <- ymd("20220331")
date2 <- mdy("03/31/2022")

# Print the dates
print(date1)
print(date2)

# Create a time object
time1 <- hm("13:45")
time2 <- hms("13:45:15")

# Print the times
print(time1)
print(time2)

# Combine the date and time objects
datetime1 <- ymd_hm("2022-03-31 13:45")
datetime2 <- mdy_hms("03/31/2022 01:45:15 PM")

# Print the datetimes
print(datetime1)
print(datetime2)

# Extract the components of a datetime object
year(datetime1)
month(datetime1)
day(datetime1)
hour(datetime1)
minute(datetime1)
second(datetime1)

# Calculate the difference between two datetime objects
time_diff <- difftime(datetime2, datetime1)
print(time_diff)

```

## 




##### Additional Resources
[Data Wrangling Cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)
