---
title: "Module 3 - Intro to Data Wrangling with Base R and dplyr"
author: "Lansing Perng"
date: "Fall 2023"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

In our intro to data wrangling module, we will cover the following topics:
- Importing data into R
- Data exploration (str(), unique(), which(), hist(), boxplot())
- Data transformations
- Subsetting data and operators in R (arithmetic, logical)
- Editing entries in a dataframe (df$col[row],df[row,col])
- Dealing with imperfect data (NAs, outliers)

```{r, message=FALSE, warning=FALSE}
?require

PKG <- c('tidyverse','here')

for (p in PKG) {
  if(!require(p,character.only = TRUE)) {  
    install.packages(p)
    require(p,character.only = TRUE)}
}
```


## Importing data
```{r}
write.csv(iris,here('data','iris.csv'), row.names = F) # we will go over writing files at the end

iris<-read.csv(here('data','iris.csv')) # import data directly from your data folder

```


## Data exploration

The first thing you should do when you're starting to work with a dataset is data exploration. We want to see what the data look like and explore the spread/distribution.
```{r}
plot(iris$Petal.Length,iris$Sepal.Length) # simple plot to visualize the data

# frequency plot using hist()
hist(iris$Sepal.Length, breaks = 10,  # breaks determines how you want counts to be grouped
        main = "Counts of Sepal Length")


# shows spread of data in summary terms using boxplot
boxplot(iris$Sepal.Length,     
        main = "Distribution of Sepal Length") 
```


## Data transformation

## Indexing and subsetting data structures

### R Operators

You can manipulate data in R using various operators, including arithmetic operators for performing calculations, and logical operators for subsetting data. A brief summary is here: <https://www.statmethods.net/management/operators.html>

## Random sampling and data generation

```{r}
set.seed(99) #will need to run a set.seet() before each randomizing line of code
x <- rnorm(50, mean = 100, sd = 25)  # create example vector

x_rsample <- sample(x, size = 30)  # randomly sample 30 points from vector
x_randomized <- sample(x)  # randomize the order of the elements in the vector
```


##### Additional Resources



