---
title: "Module 1 - Introduction to R & RStudio"
author: "Lansing Perng"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

This introductory module will introduce you to R software, the RStudio interface and cover the basics of starting to code in R. R is a powerful open source software that is free and has the flexibility to do anything you could possibly want with data. Open source means that R software is always under continuous development by a global community of scientists. People are always writing new packages to optimize data analysis and workflow all the time and R allows you to have access to them for free with a single keystroke. Its accessibility also means there are endless resources available for finding help with coding and solving statistical problems. The best part, it's 100% **FREE**! RStudio is a separate program which provides a user-friendly interface for running R code. R will run with out RStudio, but RStudio will not run without R.

## Installing R and RStudio

First, [download R & RStudio](https://posit.co/download/rstudio-desktop/) for your operating system and install it. Once you have both programs, you only need to open RStudio to run R. 

## Creating and navigating R Projects

One of the best practices when working with R is to organize your work into projects. Create a new project now by clicking on the down arrow in the upper right of your interface where it says **Project: (None)** and selecting **New Project**. Select **New Directory**&rarr;**New Project**, then name your project (e.g. 'Intro to R'). This will create a folder named 'Intro to R' in the directory you chose and store all the project files inside. 

Using R projects keeps your code organized and optimizes your R workflow. It is also helpful for code sharing and reproducibility. Working in a project sets the `working directory` to your project directory by default. Your wd is the default folder where your scripts will read in data from and output dataframes and plots to.

```{r, eval=F}
# check working directory
getwd() # your working directory which will be the project folder you just created
```

Keeping all your project files in the same place optimizes importing data and exporting outputs, making your coding experience more efficient and organized. To navigate between different projects, click on the "Projects" tab in the RStudio pane on the right side of the screen. 

Keep your projects organized by always having _four basic folders_ in your project folder:
```{r, eval=FALSE}
install.packages("here", repos = "http://cran.us.r-project.org") #usually just needs the package name argument, but needed the second argument to knit the Rmd file
library(here) ## DONT put libraries in the middle of your code!
#this is here only for the purposes of showing the importance of generating these four folders

## create basic folders for every project
folders<- c("scripts", "data", "outputs", "figures") # create vector with what you want to name the four folders

dir<-here() #assign directory to here, which is the highest level directory in this project 

create.folders <- function(folder) { # function for inserting all folders
  dir.create(file.path(dir,folder)) # we will cover function writing in a later module
}
sapply(folders,create.folders) # applies the function we just wrote to each item in the vector 'folders'
```

You can also just open the project folder and create the folders one by one. But once you have this code, it will be quicker just to copy and paste it whenever you create a new project and run it.

## Best practices in R

When writing R code, there are several other best practices to follow:

1. _Keep your environment clear_ using `rm(list = ls())`. HOWEVER, be warned that if you really start to do a lot of work in R, this is actually not a great line of code to include in your scripts. If you share code and someone else runs your entire script, they will lose all the objects in their environment and they may not want that.

2. _Avoid creating variables or objects with the same name_. The most recently run code will overwrite the previous and you will no longer have that information.

3. _Always load all libraries in the top_ section of each script.Users will know exactly what to expect and what packages are required for the script, and it avoids any possibility of trying to call a function in a library that isn't loaded, as long as you always run this chunk first.

5. _Comment your code_ using `#` to explain what it does and to help others understand your work. This is vital for data transparency, collaboration, as well as for reminding yourself how you approached an analysis if you come back to it later. You've probably noticed in the above code chunks that I've written brief descriptions of the code. Commenting is also super useful for when you're testing different scripts. You can comment out some lines so that they won't run, but leave them in your script for reference.

```{r, eval=FALSE}

rm(list = ls()) #clears your global environment (top right quadrant)
# can have this line at the top of scripts to clean environment before starting a new workflow, but should remove if you plan on sharing the code

# these hashtags allow you to describe what different lines and sections of code are doing
# if i am not working in Rmd, I also like to use these to create code sections like so:

#####################################################################################
########################## Section X: Commenting Your Code ##########################
#####################################################################################
# i used the above for section headers
val <- 2+2 # sample code to fill in the section
#val <- 3+3 # commented out because it is named the same as the above object, so will overwrite it if both are run, but i can remove the # to run it whenever

##############################################
# and i use shorter # lines for splitting subsections
```

When you type more than four # in a row, it becomes an **expandable and collapsible** section break. Click the downward arrow next to the bottom line (3rd) of the section header! When you start to write long scripts, these section breaks and the collapse functionality will be very useful. 

## R Syntax caveats and general tips

- R is **case-sensitive**!! One tiny typo can cause an error because R will not understand what you are trying to do. You have to be meticulous when you write your code, not only making sure you use the right case, but also making sure every punctuation is where it needs to be.
- **<-** means **'assign to'** and is likely the most commonly used operator when writing an R script. You can use it to assign R objects like variables, vectors, data structures (dataframes), set values, and more.
- When you assign an object, it will not display unless you **call the object** by typing out what you named it (see 'val' in above code chunk)
- `'Single quotes'` are the same as `"double quotes"` as long as you open and close with the same
- `::` is used to call a function from a specific package without loading the entire package
- **Select** an entire line easily by clicking in the left margin, where the line numbers are.
- Operators that **wrap** words/phrases can be easily applied by selecting the word/phrase and clicking the operator once.
- Use the **TAB** button liberally to complete your code, such as functions, paths,object names. Pressing TAB when you've opened a function (before you type anything within the parentheses) will also show all the available arguments and you can hover your cursor over them to 
- Use **FIND** (the magnifying glass above your script) to easily search within your script and replace a certain word/phrase.
- Some functions, such as `rm(list = ls())` and `getwd()`, are purposefully left blank. This just means there are *no required arguments*. In R, an `argument` is a value or expression the user specifies for a given function within () that tell the function what you want it to do. Some functions may require only one (or even no) arguments, whereas more complex functions will require more and can also accept many added optional arguments. 
- **Numerical** values can be written straight as numbers: `2000`, but **characters** (words and phrases) have to be written as `strings`, which are enclosed in 'quotes'

```{r}
val1 <- 2000 # will save as numerical
year <- '2000'# will save as character

#check with str()
str(val1) # str() inspects the structure of an object
str(year) # it provides objects types, dimensions, and contents
```

## R libraries
R `libraries` are collections of R functions, data, and documentation that extend the capabilities of R, also called `packages.` You can install libraries from the CRAN repository using the `install.packages()` function. You can also install libraries from source code under development using the devtools library. [Here](https://support.posit.co/hc/en-us/articles/201057987-Quick-list-of-useful-R-packages) is a list of useful packages, many of which we will cover in this course.

```{r, eval=FALSE}
# Install devtools from CRAN
install.packages("devtools", repos = "http://cran.us.r-project.org")

# Or the development version from GitHub:
# install.packages("devtools")
devtools::install_github("r-lib/devtools") # did not need to library(devtools)
#useful if you don't need to use the package repeatedly within your script
```

## R Markdown
R Markdown is a tool that allows you to create documents that combine R code, results, and text in a single file.

```{r, eval=FALSE}
# install R Markdown
install.packages("rmarkdown")
```

To create an R Markdown file, click the **New File** button in the toolbar and select **R Markdown**. Title it and write your name as the author. You can use Rmd to combine text, code, and code outputs, such as tables and plots, in a single, clean document. Rmd provides a way to share code, analysis, and results in a reproducible and transparent way.

### How to write in Rmd

Note that the below chunks is not written in typical R code convention because the chunk serves the purpose of displaying operators used in Rmd aesthetics what would not show up if I write them in the Rmd text.

1. **Headers**: within the white background in an Rmd script, `#` signify headers of different levels depending on the number of #.
```{r, eval=FALSE}
# First-level header (Title)
## Second-level header #subheader
### Third-level header #sub sub
```

Notice that you can also collapse headers in Rmd!

2. Display **code** (e.g. functions) in grey box using backticks: ```
```{r, eval=FALSE}
  `str()`. These are different from 'single quotes'.
```

3. Display **bold** text and *italics*:
```{r, eval=FALSE}
  **bold** or *italics* 
```

4. Display subscripts H~2~O and superscripts R^2^ :
```{r, eval=FALSE}
  H~2~O or R^2^ 
```

5. Write **equations** in LaTeX syntax, enclosed by `$$`:
$$E=mc^2$$
```{r, eval=FALSE}
  $$E=mc^2$$
```

6. Embed **hyperlinks** 
```{r, eval=FALSE}
[Google](https://www.google.com/). # [text to display](url)
<https://www.google.com/>. # simply display the url
```

7. **Bullets and numbering** with dashes and numbers:
```{r, eval=FALSE}
- This is a bullet point # can also use * or +
- Another bullet point
  - a sub bullet
    - sub sub bullet

1. This is a numbered item
2. Another numbered item 
  - sub item 
```

8. Add a linked table of contents by changing the 'output' section of your header to:

```{r, eval=FALSE}
output: 
  html_document:
    toc: true
# the indentations need to be exactly this way for the code to work properly!
```

9. 8. Insert **code chunks** by pressing the green `+c` button at the top right of your script window and selecting 'R'. Within the curly brackets at the top of the code chunk, you can specify how the code chunk outputs information using additional arguments after 'r':

- `echo`: A _logical_ value indicating whether the code should be **printed** in the document.
- `eval`: A _logical_ value indicating whether the code should be **executed**.
- `results`: A _character_ vector indicating what type of output should be included in the document. Possible values are "asis", "hide", "hold", "markdown", "message", "output", and "verbatim". The default is "markup", which prints text output as Markdown.
- `message`, `warning`, `error`: _Logical_ values indicating whether messages, warnings, or errors generated by the code should be printed in the document.

Defaults of all logicals are `TRUE` (T also accepted), which means that all outputs are printed by default unless otherwise specified. These arguments can be useful for making the final document look cleaner if there are outputs the reader doesn't need to see.

For example, ```{r, eval=F} will print your code chunk, but will not run it.

If you only want to hide the out put of a single line within your code chunk, you can use the `invisible` function:
```{r}
3+3 # prints output by default
invisible(3+3) # will run the code but will not display output
```

When we are finished, we can knit the Rmd document into an easily readable output document by pressing the `Knit`button above the Rmd window. When you knit the Rmd, the code gets run in a clean environment (with nothing loaded and not based in your R software). So any `install.packages()` calls may require you to add a second argument that tells it what CRAN mirror to install the package from (above in the packages section). This is usually set by default within the version of R you downloaded.

## Module Assignment

For this module, your assignment will be to produce an Rmd document that provides: 

1) <span style="color: cornflowerblue">(50 pts)</span> A **quick summary** of best practices in r in a bulleted list. 

2) <span style="color: cornflowerblue">(50 pts)</span> A description of your **expectations** for this course. Briefly describe your current level of proficiency in R and outline what you hope to get out of this course, what skills you hope to gain, what types of coding/analyses you hope to learn, and how it compares with the curriculum that is currently laid out for the course. 

3) A **reflection** of our first introductory module. Please rate the difficulty of this module from 1-10 and describe why. Was it a useful/informative module? Was it too challenging, just challenging enough, or too simple? Was it too long or too short? If you have any, please provide suggestions for future improvement. Is there content you feel should be removed or added, or does any section need clarification? Please include a code chunk of some codes you learned today you think will be the most useful for you. 
<br><br>The document as a whole should demonstrate your Rmd learned skills by including each of **9** the Rmd writing topics covered today. If it doesn't fit naturally within the flow of your paragraph, add a demo section after your *summary of best practices*. For example, demo writing an **equation** by finding a well known relationship of interest online (e.g., population growth, water balance).

I want to build an R curriculum that is geared toward student learning and success, so please be honest in your reflections so that I can make future modules as conducive to learning as possible! Each of your module assignments for the rest of the course will include 3) as a reflection section at the end of the document you hand in. I will read everyone's reflections carefully and take them into consideration.


## Getting help in R
Last but absolutely not least, there is a large and active community of R users. You can find help on R-related topics in online forums, blogs, personal githubs, etc. Additionally, there are many resources available online, including tutorials, documentation, and video courses. When in doubt, don't be afraid to ask for help! Google can be your best friend!

##### Additional Resources
- [Best Practices](https://towardsdatascience.com/best-practices-for-r-programming-ec0754010b5a)
- [R Markdown Cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf)
- [R Markdown Guide](https://www.dataquest.io/blog/r-markdown-guide-cheatsheet/)
- [Equations in R Markdown](https://rmd4sci.njtierney.com/math) (can also click through other tabs for lots of useful Rmd info!)




