---
title: "Module 10 - Advanced ggplot2"
author: "Lansing Perng"
date: "Fall 2023"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

## Get your plots publication ready

You may want to reorder the levels of one of your categorical variables. The default order is alphabetical, but something else might be more meaningful for your research.

```{r}

```


It may also be cleaner for your plot to only color the vital elements. A transparent background can come in handy for making present
```{r}
g7 <- g6 +
  theme(  panel.grid.major = element_blank(), #delete major grid lines
          panel.grid.minor = element_blank(), #delete minor grid lines
          panel.background = element_rect(fill = "transparent",color = 'black', size = 1), 
          # transparent plot area, black axis lines of thickness 1
          
          plot.background = element_rect(fill = "transparent",color = NA), # transparent page background
          plot.margin = unit(c(0.3,2,0.3,0.3), "cm")) # specify plot margins in cm
```



## ggplot legends

```{r}
      theme(  panel.grid.major = element_blank(), #delete major grid lines
              panel.grid.minor = element_blank(), #delete minor grid lines
              panel.background = element_rect(fill = "transparent",colour = 'black', size = 1),
              plot.background = element_rect(fill = "transparent",colour = NA),

              legend.position = 'right', # c(0.37,0.57),
              legend.background = element_rect(fill="transparent", color = 'black'),
              legend.key = element_rect(fill = "transparent", colour = NA),
              legend.key.size = unit(1, 'cm'),
              legend.spacing.y = unit(0.2, 'cm'),
              legend.text=element_text(size=17),
              legend.title=element_blank(),
              legend.margin = margin(2,12,8,10)) +
        guides(fill=guide_legend(byrow = T))

```



## Complex plot types
NA handling?

## Color palettes

## Counts and stacked plots

## Facets and layers

## Module Assignment
If you need help arranging the plots nicely on a page, see additional resources below.
Edit their time series graph to show anomalies instead, compared to a historical value/mean

#### Additional Resources
[Arranging plots on a page](https://ggplot2-book.org/arranging-plots.html)